---
title: "615strawberry"
author: "Yishun Zhang"
date: "2024-10-07"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Set-up

```{r}
#| label: load libraries and set options
#| warning: false
#| message: false

library(knitr)
library(kableExtra)
library(tidyverse)


```
Read the data and take a first look

```{r}
#| label: read data - glimpse 

library(tidyverse)

strawberry_data <- read_csv("C:/Users/17756/Downloads/strawberries25_v3.csv")

glimpse(strawberry_data)

head(strawberry_data)

```

I have 12699 rows and 21 columns.

All I can see from the glimpse is I have date, location, values and coefficients of variation.

Examine the data. How is it organized?

```{r}
#| label: explore organization 1 

library(tidyverse)

strawberry <- read_csv("C:/Users/17756/Downloads/strawberries25_v3.csv", col_names = TRUE)

glimpse(strawberry)

summary(strawberry)

colSums(is.na(strawberry))
str(strawberry)

distinct_states <- strawberry |> distinct(State)
print(distinct_states)

state_counts <- strawberry |> group_by(State) |> count()
print(state_counts)

summary(strawberry$Value)
hist(as.numeric(strawberry$Value), main="Distribution of Value", xlab="Value", col="skyblue", breaks=50)
state_year_check <- strawberry |> group_by(State, Year) |> count()
print(state_year_check)

duplicates_check <- strawberry |> duplicated()
sum(duplicates_check)



```

## remove columns with a single value in all rows

```{r}
#|label: function def - drop 1-item columns
library(readr)

strawberry <- read_csv("C:/Users/17756/Downloads/strawberries25_v3.csv")

glimpse(strawberry)


drop_one_value_col <- function(df) {   
  drop <- NULL  
  for(i in 1:dim(df)[2]){     
    if(n_distinct(df[[i]]) == 1){
      drop <- c(drop, i)
    }
  }
  if(!is.null(drop)){
    df <- df[,-drop]
  }
  return(df)
}

strawberry <- drop_one_value_col(strawberry)

```

To get better look at the data, look at California.

```{r}
#| label: explore California only
library(tidyverse)
library(readr)
strawberry <- read_csv("C:/Users/17756/Downloads/strawberries25_v3.csv")
glimpse(strawberry)
strawberry <- strawberry |> 
  mutate(`Data Item` = str_trim(`Data Item`, side = "both"))

strawberry <- strawberry |>
  separate(`Data Item`, into = c("Fruit", "Category", "Item", "Metric"), sep = ",", fill = "right")
strawberry
california_data <- strawberry |> filter(State == "CALIFORNIA")
glimpse(california_data)
unique(california_data$Program)
unique(california_data$Year)
unique(california_data$Category)
california_census <- california_data |> filter(Program == "CENSUS")
california_survey <- california_data |> filter(Program == "SURVEY")
glimpse(california_census)
glimpse(california_survey)


```

Explore California to understand the census and survey

```{r}
#| label: explore Calif census and survey 
library(tidyverse)
strawberry <- read_csv("C:/Users/17756/Downloads/strawberries25_v3.csv", col_names = TRUE)



california_data <- strawberry |> filter(State == "CALIFORNIA")

calif_census <- california_data |> filter(Program == "CENSUS")
calif_survey <- california_data |> filter(Program == "SURVEY")

glimpse(calif_census)
glimpse(calif_survey)
summary(calif_census)
summary(calif_survey)
drop_one_value_col <- function(df) {   
  drop <- NULL  
  for(i in 1:dim(df)[2]){     
    if(n_distinct(df[[i]]) == 1){
      drop <- c(drop, i)
    }
  }
  if(!is.null(drop)){
    df <- df[,-drop]
  }
  return(df)
}
strawberry <- strawberry |>
mutate(`Data Item` = str_replace_all(`Data Item`, " - ", ","))
#Split 'Data Item' into 4 columns
strawberry <- strawberry |>
separate_wider_delim( cols = `Data Item`,
delim = ",",
names = c("Fruit",
"Category",
"Item",
"Metric"),
too_many = "merge",
too_few = "align_start"
)
#Remove 'measured in' to metric columns
strawberry <- strawberry |>
mutate(Metric = ifelse(grepl("MEASURED IN", Item), Item, Metric), # Move the 'Item' value to 'Metric' if it contains 'MEASURED IN'
Item = ifelse(grepl("MEASURED IN", Item), NA, Item) # Set 'Item' to NA where we moved the value
)
#Remove 'production' to its correct way.
strawberry <- strawberry |>
mutate(
Item = ifelse(grepl("PRODUCTION", Metric), "PRODUCTION", Item), # Move 'PRODUCTION' to 'Item'
Metric = ifelse(grepl("PRODUCTION", Metric), sub("PRODUCTION, ", "", Metric), Metric) # Remove 'PRODUCTION' from 'Metric'
)
#Remove 'utilized' from category to Item
3
strawberry <- strawberry |>
mutate(
Item = ifelse(grepl("UTILIZED", Category, ignore.case = TRUE),
paste("UTILIZED", Item, sep = " "), # Combine 'Item' with 'Utilized'
Item), # Keep 'Item' unchanged if 'Utilized' not found
Category = ifelse(grepl("UTILIZED", Category, ignore.case = TRUE), NA, Category)# Set 'Category' to NA where 'Utilized' is moved
)
#Consider a better waty to move items in one step.
movingitem<- c("ACRES BEARING", "ACRES NON-BEARING", "ACRES GROWN", "OPERATIONS WITH AREA BEARING", "YIELD", "ACRES HARVESTED", "ACRES PLANTED", "OPERATIONS WITH AREA GROWN", "OPERATIONS WITH AREA NON-BEARING","PRODUCTION")
# Move terms from 'Metric' or 'Category' to 'Item' without replacing 'Metric' data
strawberry <- strawberry |>
mutate(Item = ifelse(grepl(paste(movingitem, collapse = "|"), Category,
ignore.case = TRUE) & is.na(Item), Category,
ifelse(grepl(paste(movingitem, collapse = "|"), Category, ignore.case = TRUE),
paste(Item, Category, sep = ", "), Item)
),
Category = ifelse(grepl(paste(movingitem, collapse = "|"), Category,
ignore.case = TRUE),
NA, Category)
)




```

### `Data Item` into (fruit, category, item)

```{r}
#|label: split Data Item
library(tidyverse)

strawberry <- read_csv("C:/Users/17756/Downloads/strawberries25_v3.csv", col_names = TRUE)

census_data <- strawberry |> filter(Program == "CENSUS")
survey_data <- strawberry |> filter(Program == "SURVEY")
drop_one_value_col <- function(df) {   
  drop <- NULL  
  for(i in 1:dim(df)[2]){     
    if(n_distinct(df[[i]]) == 1){
      drop <- c(drop, i)
    }
  }
  if(!is.null(drop)){
    df <- df[,-drop]
  }
  return(df)
}
census_data <- drop_one_value_col(census_data)
survey_data <- drop_one_value_col(survey_data)

```


## now exam the Fruit column -- find hidden sub-columns

```{r}

library(tidyverse)

strawberry <- read_csv("C:/Users/17756/Downloads/strawberries25_v3.csv", col_names = TRUE)

strawberry <- strawberry |>
separate_wider_delim(
cols = Domain,
delim = " , ",
names = c("Area Grown", "Fertilize", "Organic", "Chemical"),
too_many = "merge",
too_few = "align_start"
)
#Loading variables to each column
strawberry <- strawberry |>
mutate(
Chemical = ifelse(grepl("CHEMICAL", `Area Grown`, ignore.case = TRUE), `Area Grown`, NA),
Organic = ifelse(grepl("ORGANIC", `Area Grown`, ignore.case = TRUE), `Area Grown`, NA),
Fertilize = ifelse(grepl("FERTILIZER", `Area Grown`, ignore.case = TRUE), `Area Grown`, NA),
`Area Grown`= ifelse(grepl("CHEMICAL|ORGANIC|FERTILIZER", `Area Grown`, ignore.case = TRUE), NA, `Area Grown`)
)
#Dealing with 'Domain Category' column
strawberry <- strawberry |>
mutate(
Chemical = ifelse(grepl("CHEMICAL", `Domain Category`, ignore.case = TRUE),
`Domain Category`,
Chemical),
Organic = ifelse(grepl("ORGANIC", `Domain Category`, ignore.case = TRUE),
`Domain Category`,
Organic),
Fertilize = ifelse(grepl("FERTILIZER", `Domain Category`, ignore.case = TRUE),
`Domain Category`,
Fertilize),
`Area Grown` = ifelse(grepl("AREA", `Domain Category`, ignore.case = TRUE),
`Domain Category`,
`Area Grown`),
`Domain Category` = ifelse(grepl("CHEMICAL|ORGANIC|FERTILIZER|AREA", `Domain Category`, ignore.case = TRUE), NA, `Domain Category`)
)
#Move 'Total' to its best place
strawberry <- strawberry |> 
  mutate(`Data Item` = str_trim(`Data Item`, side = "both"))

strawberry <- strawberry |>
  separate(`Data Item`, into = c("Fruit", "Category", "Item", "Metric"), sep = ",", fill = "right")
strawberry
strawberry <- strawberry |>
mutate(Item = ifelse(grepl("Total", `Area Grown`, ignore.case = TRUE),
paste("Total", Item, sep = " "),
Item),
`Area Grown` = ifelse(grepl("Total", `Area Grown`, ignore.case = TRUE), NA, `Area Grown`)
)



  
```

```{r}


library(tidyverse)

strawberry <- read_csv("C:/Users/17756/Downloads/strawberries25_v3.csv", col_names = TRUE)
strawberry <- strawberry |>
separate_wider_delim(
cols = Domain,
delim = " , ",
names = c("Area Grown", "Fertilize", "Organic", "Chemical"),
too_many = "merge",
too_few = "align_start"
)
#Loading variables to each column
strawberry <- strawberry |>
mutate(
Chemical = ifelse(grepl("CHEMICAL", `Area Grown`, ignore.case = TRUE), `Area Grown`, NA),
Organic = ifelse(grepl("ORGANIC", `Area Grown`, ignore.case = TRUE), `Area Grown`, NA),
Fertilize = ifelse(grepl("FERTILIZER", `Area Grown`, ignore.case = TRUE), `Area Grown`, NA),
`Area Grown`= ifelse(grepl("CHEMICAL|ORGANIC|FERTILIZER", `Area Grown`, ignore.case = TRUE), NA, `Area Grown`)
)
#Dealing with 'Domain Category' column
strawberry <- strawberry |>
mutate(
Chemical = ifelse(grepl("CHEMICAL", `Domain Category`, ignore.case = TRUE),
`Domain Category`,
Chemical),
Organic = ifelse(grepl("ORGANIC", `Domain Category`, ignore.case = TRUE),
`Domain Category`,
Organic),
Fertilize = ifelse(grepl("FERTILIZER", `Domain Category`, ignore.case = TRUE),
`Domain Category`,
Fertilize),
`Area Grown` = ifelse(grepl("AREA", `Domain Category`, ignore.case = TRUE),
`Domain Category`,
`Area Grown`),
`Domain Category` = ifelse(grepl("CHEMICAL|ORGANIC|FERTILIZER|AREA", `Domain Category`, ignore.case = TRUE), NA, `Domain Category`)
)
strawberry <- strawberry |>
mutate(Chemical = str_replace_all(Chemical, "[,:=()]", ","))
#Split it into three columns
strawberryc<- strawberry |>
separate_wider_delim(
cols = Chemical,
delim = ",",
names = c("Type", "Ingredient", "Code"), #Separate Chemical into type, ingredient, and code.
too_many = "merge",
too_few = "align_start"
)
#Filling in the columns
strawberryc<- strawberryc |>
mutate(
Type = ifelse(Type == "CHEMICAL" | is.na(Type), Ingredient, Type),Ingredient = ifelse(!is.na(Ingredient), str_extract(Code, "\\b[A-Za-z\\-\\.\\s]+\\b"), Ingredient), 
Code = str_replace(Code, "\\b[A-Za-z\\-\\.\\s]+\\b", "")
)
#Clean 'Code' Column
strawberryc <- strawberryc |>
mutate(
Code = str_replace_all(Code, "ˆ\\s*,+|,+\\s*$|\\s*,\\s*,+", ""),
Code = str_trim(Code)
)
head(strawberryc)

```
```{r}
# Load necessary libraries
library(tidyverse)

# Read the dataset
strawberry_data <- read_csv("C:/Users/17756/Downloads/strawberries25_v3.csv")

# Convert 'Value' column to numeric, replacing non-numeric placeholders with NA
strawberry_data <- strawberry_data %>%
  mutate(Value = as.numeric(Value))

# Summary statistics of numeric columns
numeric_summary <- strawberry_data %>% 
  summarise(across(where(is.numeric), list(mean = mean, sd = sd, min = min, max = max), na.rm = TRUE))
print(numeric_summary)

# Count unique states and programs
state_counts <- strawberry_data %>% count(State)
program_counts <- strawberry_data %>% count(Program)

# Distribution of 'Value' column
ggplot(strawberry_data, aes(x = Value)) +
  geom_histogram(bins = 50, fill = "skyblue", color = "black") +
  labs(title = 'Distribution of "Value" Column', x = 'Value', y = 'Frequency') +
  theme_minimal()

# Top 10 states with the most entries
top_states <- state_counts %>%
  arrange(desc(n)) %>%
  head(10)

ggplot(top_states, aes(x = reorder(State, n), y = n)) +
  geom_bar(stat = "identity", fill = "orange", color = "black") +
  coord_flip() +
  labs(title = "Top 10 States with Most Entries", x = "State", y = "Count") +
  theme_minimal()

# Year-wise data distribution
year_counts <- strawberry_data %>% 
  count(Year) %>%
  arrange(Year)

ggplot(year_counts, aes(x = as.factor(Year), y = n)) +
  geom_bar(stat = "identity", fill = "green", color = "black") +
  labs(title = "Data Entries by Year", x = "Year", y = "Count") +
  theme_minimal()

# Correlation matrix for numeric columns
numeric_cols <- strawberry_data %>% select(where(is.numeric))
correlation_matrix <- cor(numeric_cols, use = "complete.obs")


# Output summaries
print("Summary of Numeric Data:")
print(numeric_summary)

print("Top 10 States by Entry Count:")
print(top_states)

```
EDA analysis:
Summary Statistics of Numeric Data
We calculated the mean, standard deviation, minimum, and maximum for numeric columns like Value.
Observations:
1.The range of values in the Value column might be large, with the mean and median providing a sense of central tendency.
2.A high standard deviation suggests a wide spread of data, whereas a low standard deviation indicates the data is more concentrated.

Distribution of the "Value" Column
The histogram shows the frequency distribution of the Value column:
1.A right-skewed distribution indicates most data points have smaller values, with some large outliers.
2.A left-skewed distribution suggests most data points have larger values, with some smaller outliers.
Potential findings:
1.A flat or multi-modal distribution may indicate the need for subgroup analysis based on states, years, or other factors.
2.For unusual distributions, consider data transformation, such as log transformation or normalization.

Top 10 States with the Most Data Entries
The bar chart highlights the top 10 states with the most entries:
1.California has the highest number of data entries, significantly more than others.
Florida is the second-largest contributor.
2.Other states contribute fewer records and may have less influence on the overall analysis.
Analysis & Recommendations:
1.Focus on California and Florida for detailed trend and characteristic analyses, as they dominate the dataset.
2.For states with fewer records, consider aggregating data for regional comparisons.

Yearly Data Distribution
The bar chart shows the distribution of data entries across different years:
1.A high number of entries in certain years might indicate more detailed surveys conducted during those times.
2.Low or missing data in specific years should be further investigated.
Findings:
1.If the data distribution across years is even, it is well-suited for trend analysis.
2.If uneven, consider standardizing the data volume across years for better comparisons.
